---
title: "Practical Machine Learning Project"
author: "Sreekantha Thimmareddy"
date: "Sunday, April 26, 2015"
output: html_document
---

Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

Data 
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

What to submit

The goal of the project is to predict the manner in which they did the exercise. 

1. submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5.

2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading.

For this assignment I analyzed the provided data to determine what activity an individual perform. To do this I made use of caret and randomForest, this allowed me to generate correct answers for each of the 20 test data cases provided in this assignment. I made use of a seed value for consistent results.


```{r, echo=FALSE, results='hide',message=FALSE,error=TRUE}
library(data.table)
library(caret)
library(ggplot2)
library(knitr)
library(xtable)
library(randomForest)
library(RCurl)
library(Hmisc)
library(foreach)
library(doParallel)
library(gridExtra)
options(warn=-1)
set.seed(2048)
```

##Data Processing

First, I loaded the data both from the provided training and test data provided by COURSERA. Some values contained a "#DIV/0!" that I replaced with an NA value. I also casted all columns 8 to the end to be numeric.


```{r}
URLTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
t1 <- getURL(URLTrain)

URLEvaluate <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
e1 <- getURL(URLEvaluate)


training_data <- read.csv(textConnection(t1), na.strings=c("#DIV/0!")) 

evaluation_data <- read.csv(textConnection(e1), na.strings=c("#DIV/0!")) 


for(i in c(8:ncol(training_data)-1)) {training_data[,i] = as.numeric(as.character(training_data[,i]))}

for(i in c(8:ncol(evaluation_data)-1)) {evaluation_data[,i] = as.numeric(as.character(evaluation_data[,i]))}
```


Some columns were mostly blank. These did not contribute well to the prediction. I chose a feature set that only included complete columns. We also remove user name, timestamps and windows.

Determine and display out feature set.

```{r}
feature_set <- colnames(training_data[colSums(is.na(training_data)) == 0])[-(1:7)]
model_data <- training_data[feature_set]
feature_set


idx <- createDataPartition(y=model_data$classe, p=0.60, list=FALSE )

training <- model_data[idx,]
testing <- model_data[-idx,]
```

#Exploratory data analysis

Exploratory data analysis showed multiple class-based clusters. 
Based on clusters distribution I decided to use random forests

```{r, echo=FALSE}
grid.arrange(
  ggplot(training, aes(x=accel_arm_z, y=yaw_belt, colour=classe)) + geom_point(),
  ggplot(training, aes(x=pitch_dumbbell, y=total_accel_belt, colour=classe)) + geom_point(),
  ggplot(training, aes(x=total_accel_arm, y=yaw_forearm, colour=classe)) + geom_point(),
  ggplot(training, aes(x=roll_arm, y=total_accel_forearm, colour=classe)) + geom_point(),
  ncol=2
)
```

##Model building and testing

We now build 5 random forests with 150 trees each. We make use of parallel processing to build this model. I found several examples of how to perform parallel processing with random forests in R, this provided a great speedup.
Provide error reports for both training and test data.

```{r}
registerDoParallel()
x <- training[-ncol(training)]
y <- training$classe

rf <- foreach(ntree=rep(150, 6), .combine=randomForest::combine, .packages='randomForest') %dopar% {
  randomForest(x, y, ntree=ntree) 
}


predictions1 <- predict(rf, newdata=training)
confusionMatrix(predictions1,training$classe)


predictions2 <- predict(rf, newdata=testing)
confusionMatrix(predictions2,testing$classe)
```

##Conclusions and Test Data Submit

As can be seen from the confusion matrix this model is very accurate. I did experiment with PCA and other models, but did not get as good of accuracy. Because my test data was around 99% accurate I expected nearly all of the submitted test cases to be correct. It turned out they were all correct.

Prepare the submission. (using COURSERA provided code)

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}


x <- evaluation_data
x <- x[feature_set[feature_set!='classe']]
answers <- predict(rf, newdata=x)

answers

pml_write_files(answers)
```




